Is this normal? A new projection pursuit index to assess a sample against a multivariate null distribution

### 2024-07-24 paper.qmd
References fixed

---
title: "Is this normal? A new projection pursuit index to assess a sample against a multivariate null distribution"
author: 
  - name: Annalisa Calvi
    affiliations:
      name: Monash University
      department: School of Mathematics
    email: annalisa.calvi@monash.edu
    ORCID: 0009-0000-2977-0074
  - name: Ursula Laa
    affiliations:
      name: University of Natural Resources and Life Sciences, Vienna
      department: Institute of Statistics
    email: ursula.laa@boku.ac.at
    ORCID: 0000-0002-0249-6439
  - name: Dianne Cook
    affiliations:
      name: Monash University
      department: Department of Econometrics and Business Statistics
    email: dicook@monash.edu
    ORCID: 0000-0002-3813-7155
format:
  tandf-pdf:
    keep-tex: true  
    linestretch: 2
    fontsize: 12pt
abstract: |
  Many data problems contain some reference or normal conditions, upon which to compare newly collected data. This scenario occurs in data collected as part of clinical trials to detect adverse events, and or for measuring climate change against historical norms. The data is typically multivariate, and often the normal ranges are specified by a multivariate normal distribution. The work presented in this paper develops methods to compare the new sample against the reference distribution with high-dimensional visualisation. It uses a projection pursuit guided tour to produce a sequence of low-dimensional projections steered towards those where the new sample is most different from the reference. A new projection pursuit index is defined, and the drawing of the projected ellipse is computed analytically. The methods are implemented in the R package, `tourr`. 
keywords:
  - exploratory data analysis
  - statistical graphics
  - dimension reduction
  - high-dimensional visualisation
  - climate change
  - clinical trials
bibliography: bibliography.bib  
header-includes: | 
  \usepackage{amsmath}
  \usepackage{amsthm}
  \usepackage{float}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{bm}
  \def\tightlist{}
  \usepackage{setspace}
  \newcommand\pD{$p\text{-}D$}
  \newcommand\kD{$k\text{-}D$}
  \newcommand\dD{$d\text{-}D$}
  \newcommand\gD{$2\text{-}D$}
  \newcommand\bD{$1\text{-}D$}
  \newtheorem*{theorem}{Theorem}
---

@@

This paper describes a new projection pursuit index which is optimized by projections where a new sample is most distant from the existing normal distribution. It is organised as follows. @sec-background provides more context for the methods and visualisation. @sec-anomaly-index provides the details of the new index, and example use is illustrated in @sec-examples.

@@

## Implementation

@@

This example is motivated by the weather data example from @filzmoser2018. We illustrate using the anomaly index to compare potential outliers with a reference normal distribution that is derived using robust methods on the original data. The data contains 16 numerical measurements that are averages across the three summer months June, July and August, reported for 68 years (1955 - 2022) (see @filzmoser2018 for more details).

@@

Here we will consider points to be outlying if they are more than $5 \sigma$ away from that mean value. With $p=16$ this corresponds to a Mahalanobis distance of $60$ or larger. This will identify $20$ out of the $68$ observations as outlying. Since these are outlying in different combinations of variables, as found in @filzmoser2018, we will further separate the outlying points into clusters based on  similarity in direction. The similarity is computed by first normalizing observations to have length $1$ and then apply k-means clustering with Euclidean distance. We use the Dunn index @dunn, computed using @fpc, to select the preferred number of clusters, $k=5$.

@@

To compare our results to what is obtained by outlier detection methods, we reproduce one of the results from @filzmoser2018 in @fig-shapley. On top we can see the cluster assignments for all points that we have identified as outliers using the $p$-dimensional Mahalanobis distance.

For example we see that cluster 4 contains the years 1962 and 1996, and for these two years the result from @filzmoser2018 also indicates unusually low values for the minimum teperature, while there is disagreement between outlyingness in other variables. Note that compared to the years 1965 and 1966 they have unusually low minimum temperature while having average values for the number of days with maximum temperature above 25°C, and this contrast is also captured in the projection.

@@

Reproduction of the results from @filzmoser2018, showing which cells have been identified as outlying by their Shapley Cell Detector algorithm. Red cells indicate values higher than expected, blue cells values lower than expected. For years that have been identified as outliers in our approach we show the cluster assignment at the top.
:::

@@

The work is related to outlier detection methods [see, @filzmoser2018, for example], particularly those that use robust statistics, as illustrated in the weather example. Methods discussed in @directional-outlyingness and @donoho1982 describe detecting outlyingness using projections of the data. When these projections are collected and used as a large set, they define a high-dimensional confidence region of an indeterminate shape. There is no analytical definition. However, we could explore the data relative to the high-dimensional shapes using the same approach described in @sec-background. You would generate points on the surface of the irregular shape, overlay the data on this to visualise with a tour.

@@

The full paper, code and data to reproduce the work, are publicly available at https://github.com/uschiLaa/anomaly_ppi.

>>>>

---
title: "Is this normal? A new projection pursuit index to assess a sample against a multivariate null distribution"
author: 
  - name: Annalisa Calvi
    affiliations:
      name: Monash University
      department: School of Mathematics
    email: annalisa.calvi@monash.edu
    ORCID: 0009-0000-2977-0074
  - name: Ursula Laa
    affiliations:
      name: University of Natural Resources and Life Sciences, Vienna
      department: Institute of Statistics
    email: ursula.laa@boku.ac.at
    ORCID: 0000-0002-0249-6439
  - name: Dianne Cook
    affiliations:
      name: Monash University
      department: Department of Econometrics and Business Statistics
    email: dicook@monash.edu
    ORCID: 0000-0002-3813-7155
format:
  tandf-pdf:
    keep-tex: true  
    linestretch: 2
    fontsize: 12pt
abstract: |
  Many data problems contain some reference or normal conditions, upon which to compare newly collected data. This scenario occurs in data collected as part of clinical trials to detect adverse events, or for measuring climate change against historical norms. The data is typically multivariate, and often the normal ranges are specified by a multivariate normal distribution. The work presented in this paper develops methods to compare the new sample against the reference distribution with high-dimensional visualisation. It uses a projection pursuit guided tour to produce a sequence of low-dimensional projections steered towards those where the new sample is most different from the reference. A new projection pursuit index is defined, and the drawing of the projected ellipse is computed analytically. The methods are implemented in the R package, `tourr`. 
keywords:
  - exploratory data analysis
  - statistical graphics
  - dimension reduction
  - high-dimensional visualisation
  - climate change
  - clinical trials
bibliography: bibliography.bib  
header-includes: | 
  \usepackage{amsmath}
  \usepackage{amsthm}
  \usepackage{float}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{bm}
  \def\tightlist{}
  \usepackage{setspace}
  \newcommand\pD{$p\text{-}D$}
  \newcommand\kD{$k\text{-}D$}
  \newcommand\dD{$d\text{-}D$}
  \newcommand\gD{$2\text{-}D$}
  \newcommand\bD{$1\text{-}D$}
  \newtheorem*{theorem}{Theorem}
---

@@

This paper describes a new projection pursuit index which is optimized by projections where a new sample is most distant from the existing normal distribution. It is organised as follows. @sec-background provides more context for the methods and visualisation. @sec-anomaly-index provides the details of the new index, and @sec-implementation describes the implementation. Example use is illustrated in @sec-examples.

@@

## Implementation {#sec-implementation}

@@

This example is motivated by the weather data example from @MAYRHOFER2023. We illustrate using the anomaly index to compare potential outliers with a reference normal distribution that is derived using robust methods on the original data. The data contains 16 numerical measurements that are averages across the three summer months June, July and August, reported for 68 years (1955 - 2022) (see @MAYRHOFER2023 for more details).

@@

Here we will consider points to be outlying if they are more than $5 \sigma$ away from that mean value. With $p=16$ this corresponds to a Mahalanobis distance of $60$ or larger. This will identify $20$ out of the $68$ observations as outlying. Since these are outlying in different combinations of variables, as found in @MAYRHOFER2023, we will further separate the outlying points into clusters based on  similarity in direction. The similarity is computed by first normalizing observations to have length $1$ and then apply k-means clustering with Euclidean distance. We use the Dunn index @dunn, computed using @fpc, to select the preferred number of clusters, $k=5$.

@@

To compare our results to what is obtained by outlier detection methods, we reproduce one of the results from @MAYRHOFER2023 in @fig-shapley. On top we can see the cluster assignments for all points that we have identified as outliers using the $p$-dimensional Mahalanobis distance.

For example we see that cluster 4 contains the years 1962 and 1996, and for these two years the result from @MAYRHOFER2023 also indicates unusually low values for the minimum teperature, while there is disagreement between outlyingness in other variables. Note that compared to the years 1965 and 1966 they have unusually low minimum temperature while having average values for the number of days with maximum temperature above 25°C, and this contrast is also captured in the projection.

@@

Reproduction of the results from @MAYRHOFER2023, showing which cells have been identified as outlying by their Shapley Cell Detector algorithm. Red cells indicate values higher than expected, blue cells values lower than expected. For years that have been identified as outliers in our approach we show the cluster assignment at the top.
:::

@@

The work is related to outlier detection methods [see, @MAYRHOFER2023, for example], particularly those that use robust statistics, as illustrated in the weather example. Methods discussed in @directional-outlyingness and @donoho1982 describe detecting outlyingness using projections of the data. When these projections are collected and used as a large set, they define a high-dimensional confidence region of an indeterminate shape. There is no analytical definition. However, we could explore the data relative to the high-dimensional shapes using the same approach described in @sec-background. You would generate points on the surface of the irregular shape, overlay the data on this to visualise with a tour.

@@

## Supplementary Materials {-}
The full paper, code and data to reproduce the work, are publicly available at https://github.com/uschiLaa/anomaly_ppi.

## References {-}

### 2024-07-24 paper.tex
References fixed

\thanks{CONTACT: Annalisa
Calvi. Email: \href{mailto:annalisa.calvi@monash.edu}{\nolinkurl{annalisa.calvi@monash.edu}}. Ursula
Laa. Email: \href{mailto:ursula.laa@boku.ac.at}{\nolinkurl{ursula.laa@boku.ac.at}}. Dianne
Cook. Email: \href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} School of Mathematics, Monash
University,  \\ \textsuperscript{2} Institute of Statistics, University
of Natural Resources and Life Sciences,
Vienna,  \\ \textsuperscript{3} Department of Econometrics and Business
Statistics, Monash University,  
\begin{abstract}
Many data problems contain some reference or normal conditions, upon
which to compare newly collected data. This scenario occurs in data
collected as part of clinical trials to detect adverse events, and or
for measuring climate change against historical norms. The data is
typically multivariate, and often the normal ranges are specified by a
multivariate normal distribution. The work presented in this paper
develops methods to compare the new sample against the reference
distribution with high-dimensional visualisation. It uses a projection
pursuit guided tour to produce a sequence of low-dimensional projections
steered towards those where the new sample is most different from the
reference. A new projection pursuit index is defined, and the drawing of
the projected ellipse is computed analytically. The methods are
implemented in the R package, \texttt{tourr}.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
exploratory data analysis\sep statistical graphics\sep dimension
reduction\sep high-dimensional visualisation\sep climate change\sep 
clinical trials
\end{keywords}

@@

More generally, projection pursuit (Friedman and Tukey 1974; Jones and
Sibson 1987; Huber 1985) defines a quantitative criterion for the
interestingness of a projection (a projection pursuit index), and
searches the space of possible projections for the most interesting one
to display. We can also define sequences of interpolated linear
projections to better understand a multivariate distribution. Animating
a randomly selected interpolated sequence of linear projections is
called a grand tour (Asimov 1985; Buja and Asimov 1986; Buja et al.
2005; D. Cook et al. 2006; Lee et al. 2022). The combination of these
two approaches would then use a projection pursuit index to select
interesting projections, but display them via an interpolated path to
provide context. This is called a guided tour (Dianne Cook et al. 1995).

@@

This paper describes a new projection pursuit index which is optimized
by projections where a new sample is most distant from the existing
normal distribution. It is organised as follows.
Section~\ref{sec-background} provides more context for the methods and
visualisation. Section~\ref{sec-anomaly-index} provides the details of
the new index, and example use is illustrated in
Section~\ref{sec-examples}.

@@

To define a measure of an interesting projection is to maximize the
average Mahahlanobis distance (Mahalanobis 1936) in the projection for a
subset of points, \(W\). The set of points could be chosen in different
ways, but the default is those that are outside the specified ellipsoid
in \pD{}. Alternatives could be to select a set of observations with the
largest Mahalanobis distance, manually select observations or possibly a
group of points identified via clustering of the extremes.

@@

\section{Implementation}\label{implementation}

This is implemented in the \texttt{tourr} (Hadley Wickham et al. 2011;
H. Wickham et al. 2024) package, where the projected ellipsoid can be
drawn for each projection. The guided tour will take arguments
specifying the data, and the null variance-covariance matrix.

@@

This example is motivated by the weather data example from Filzmoser,
Hron, and Templ (2018). We illustrate using the anomaly index to compare
potential outliers with a reference normal distribution that is derived
using robust methods on the original data. The data contains 16
numerical measurements that are averages across the three summer months
June, July and August, reported for 68 years (1955 - 2022) (see
Filzmoser, Hron, and Templ (2018) for more details).

To estimate the underlying normal distribution the data is first
centered and scaled using the median and the median absolute deviation
(MAD), before applying the minimum covariance determinant (MCD)
estimator (Peter J. Rousseeuw 1985) using the implementation in Maechler
et al. (2024). The MCD estimates for the mean vector and the
variance-covariance matrix are then used to define the reference normal
distribution.

Here we will consider points to be outlying if they are more than
\(5 \sigma\) away from that mean value. With \(p=16\) this corresponds
to a Mahalanobis distance of \(60\) or larger. This will identify \(20\)
out of the \(68\) observations as outlying. Since these are outlying in
different combinations of variables, as found in Filzmoser, Hron, and
Templ (2018), we will further separate the outlying points into clusters
based on similarity in direction. The similarity is computed by first
normalizing observations to have length \(1\) and then apply k-means
clustering with Euclidean distance. We use the Dunn index Halkidi,
Batistakis, and Vazirgiannis (2002), computed using Hennig (2024), to
select the preferred number of clusters, \(k=5\).

@@

To compare our results to what is obtained by outlier detection methods,
we reproduce one of the results from Filzmoser, Hron, and Templ (2018)
in Figure~\ref{fig-shapley}. On top we can see the cluster assignments
for all points that we have identified as outliers using the
\(p\)-dimensional Mahalanobis distance.

For example we see that cluster 4 contains the years 1962 and 1996, and
for these two years the result from Filzmoser, Hron, and Templ (2018)
also indicates unusually low values for the minimum teperature, while
there is disagreement between outlyingness in other variables. Note that
compared to the years 1965 and 1966 they have unusually low minimum
temperature while having average values for the number of days with
maximum temperature above 25°C, and this contrast is also captured in
the projection.

@@

\caption{\label{fig-shapley}Reproduction of the results from Filzmoser,
Hron, and Templ (2018), showing which cells have been identified as
outlying by their Shapley Cell Detector algorithm. Red cells indicate
values higher than expected, blue cells values lower than expected. For
years that have been identified as outliers in our approach we show the
cluster assignment at the top.}

@@

The work is related to outlier detection methods (see, Filzmoser, Hron,
and Templ 2018, for example), particularly those that use robust
statistics, as illustrated in the weather example. Methods discussed in
Peter J. Rousseeuw, Raymaekers, and Hubert (2018) and Donoho and Gasko
(1992) describe detecting outlyingness using projections of the data.
When these projections are collected and used as a large set, they
define a high-dimensional confidence region of an indeterminate shape.
There is no analytical definition. However, we could explore the data
relative to the high-dimensional shapes using the same approach
described in Section~\ref{sec-background}. You would generate points on
the surface of the irregular shape, overlay the data on this to
visualise with a tour.

@@

@@

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-As85}
Asimov, D. 1985. {``{T}he {G}rand {T}our: {A} {T}ool for {V}iewing
{M}ultidimensional {D}ata.''} \emph{SIAM Journal of Scientific and
Statistical Computing} 6 (1): 128--43.

\bibitem[\citeproctext]{ref-BA86b}
Buja, A., and D. Asimov. 1986. {``{G}rand {T}our {M}ethods: {A}n
{O}utline.''} \emph{Computing Science and Statistics} 17: 63--67.

\bibitem[\citeproctext]{ref-BCAH05}
Buja, A., D. Cook, D. Asimov, and C. Hurley. 2005. {``{C}omputational
{M}ethods for {H}igh-{D}imensional {R}otations in {D}ata
{V}isualization.''} In \emph{Handbook of Statistics: Data Mining and
Visualization}, edited by C. R. Rao, E. J. Wegman, and J. L. Solka,
391--414. Amsterdam, The Netherlands: Elsevier/North-Holland.

@@

\bibitem[\citeproctext]{ref-CLBW06}
Cook, D., E.-K. Lee, A. Buja, and H. Wickham. 2006. {``{G}rand {T}ours,
{P}rojection {P}ursuit {G}uided {T}ours and {M}anual {C}ontrols.''} In
\emph{Handbook of {D}ata {V}isualization}, edited by C.-H. Chen, W.
Härdle, and A. Unwin. Berlin: Springer.

@@

\bibitem[\citeproctext]{ref-filzmoser2018}
Filzmoser, P., K. Hron, and M Templ. 2018. \emph{Applied Compositional
Data Analysis}. Cham, Switzerland: Springer.
\url{https://doi.org/10.1007/978-3-319-96422-5}.

\bibitem[\citeproctext]{ref-FR02}
Fraley, C., and A. E. Raftery. 2002. {``Model-Based {C}lustering,
{D}iscriminant {A}nalysis, {D}ensity {E}stimation.''} \emph{Journal of
the American Statistical Association} 97: 611--31.

\bibitem[\citeproctext]{ref-FT74}
Friedman, J. H., and J. W. Tukey. 1974. {``{A} {P}rojection {P}ursuit
{A}lgorithm for {E}xploratory {D}ata {A}nalysis.''} \emph{IEEE
Transactions on Computing C} 23: 881--89.

\bibitem[\citeproctext]{ref-Ga71}
Gabriel, K. R. 1971. {``The {B}iplot {G}raphical {D}isplay of {M}atrices
with {A}pplications to {P}rincipal {C}omponent {A}nalysis.''}
\emph{Biometrika} 58: 453--67.

\bibitem[\citeproctext]{ref-GH96}
Gower, J. C., and D. J. Hand. 1996. \emph{Biplots}. London: Chapman;
Hall.

@@

\bibitem[\citeproctext]{ref-Hu85}
Huber, P. J. 1985. {``{P}rojection {P}ursuit (with Discussion).''}
\emph{Annals of Statistics} 13: 435--525.

\bibitem[\citeproctext]{ref-JS87}
Jones, M. C., and R. Sibson. 1987. {``{W}hat Is {P}rojection {P}ursuit?
(With Discussion).''} \emph{Journal of the Royal Statistical Society,
Series A} 150: 1--36.

\bibitem[\citeproctext]{ref-lib-med-liver-norms}
Lala, V., M. Zubair, and D. A. Minter. 2023. {``Liver Function Tests.''}
\url{https://www.ncbi.nlm.nih.gov/books/NBK482489/}.

@@

\bibitem[\citeproctext]{ref-robustbase}
Maechler, Martin, Peter Rousseeuw, Christophe Croux, Valentin Todorov,
Andreas Ruckstuhl, Matias Salibian-Barrera, Tobias Verbeke, Manuel
Koller, Eduardo L. T. Conceicao, and Maria Anna di Palma. 2024.
\emph{Robustbase: Basic Robust Statistics}.
\url{http://robustbase.r-forge.r-project.org/}.

\bibitem[\citeproctext]{ref-mahalanobis}
Mahalanobis, P. C. 1936. {``On the Generalised Distance in
Statistics.''} \emph{Sankhya A} 80 (Suppl 1): 1--7.
\url{https://doi.org/10.1007/s13171-019-00164-5}.

@@

\bibitem[\citeproctext]{ref-tourr-cran}
Wickham, H., D. Cook, N. Spyrison, N. Laa, H. S. Zhang, and S. Lee.
2024. \emph{Tour Methods for Multivariate Data Visualisation}.
\url{https://CRAN.R-project.org/package=tourr}.

>>>>

\thanks{CONTACT: Annalisa
Calvi. Email: \href{mailto:annalisa.calvi@monash.edu}{\nolinkurl{annalisa.calvi@monash.edu}}. Ursula
Laa. Email: \href{mailto:ursula.laa@boku.ac.at}{\nolinkurl{ursula.laa@boku.ac.at}}. Dianne
Cook. Email: \href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} School of Mathematics, Monash
University,  \\ \textsuperscript{2} Institute of Statistics, University
of Natural Resources and Life Sciences,
Vienna,  \\ \textsuperscript{3} Department of Econometrics and Business
Statistics, Monash University,  
\begin{abstract}
Many data problems contain some reference or normal conditions, upon
which to compare newly collected data. This scenario occurs in data
collected as part of clinical trials to detect adverse events, or for
measuring climate change against historical norms. The data is typically
multivariate, and often the normal ranges are specified by a
multivariate normal distribution. The work presented in this paper
develops methods to compare the new sample against the reference
distribution with high-dimensional visualisation. It uses a projection
pursuit guided tour to produce a sequence of low-dimensional projections
steered towards those where the new sample is most different from the
reference. A new projection pursuit index is defined, and the drawing of
the projected ellipse is computed analytically. The methods are
implemented in the R package, \texttt{tourr}.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
exploratory data analysis\sep statistical graphics\sep dimension
reduction\sep high-dimensional visualisation\sep climate change\sep 
clinical trials
\end{keywords}

@@

More generally, projection pursuit (Friedman and Tukey 1974; Jones and
Sibson 1987; Huber 1985) defines a quantitative criterion for the
interestingness of a projection (a projection pursuit index), and
searches the space of possible projections for the most interesting one
to display. We can also define sequences of interpolated linear
projections to better understand a multivariate distribution. Animating
a randomly selected interpolated sequence of linear projections is
called a grand tour (Asimov 1985; Buja and Asimov 1986; Buja et al.
2005; Cook et al. 2006; Lee et al. 2022). The combination of these two
approaches would then use a projection pursuit index to select
interesting projections, but display them via an interpolated path to
provide context. This is called a guided tour (Cook et al. 1995).

@@

This paper describes a new projection pursuit index which is optimized
by projections where a new sample is most distant from the existing
normal distribution. It is organised as follows.
Section~\ref{sec-background} provides more context for the methods and
visualisation. Section~\ref{sec-anomaly-index} provides the details of
the new index, and Section~\ref{sec-implementation} describes the
implementation. Example use is illustrated in
Section~\ref{sec-examples}.

@@

To define a measure of an interesting projection is to maximize the
average Mahalanobis distance (Mahalanobis 1936) in the projection for a
subset of points, \(W\). The set of points could be chosen in different
ways, but the default is those that are outside the specified ellipsoid
in \pD{}. Alternatives could be to select a set of observations with the
largest Mahalanobis distance, manually select observations or possibly a
group of points identified via clustering of the extremes.

@@

\section{Implementation}\label{sec-implementation}

This is implemented in the \texttt{tourr} (Wickham et al. 2011, 2024)
package, where the projected ellipsoid can be drawn for each projection.
The guided tour will take arguments specifying the data, and the null
variance-covariance matrix.

@@

This example is motivated by the weather data example from Mayrhofer and
Filzmoser (2023). We illustrate using the anomaly index to compare
potential outliers with a reference normal distribution that is derived
using robust methods on the original data. The data contains 16
numerical measurements that are averages across the three summer months
June, July and August, reported for 68 years (1955 - 2022) (see
Mayrhofer and Filzmoser (2023) for more details).

To estimate the underlying normal distribution the data is first
centered and scaled using the median and the median absolute deviation
(MAD), before applying the minimum covariance determinant (MCD)
estimator (Rousseeuw 1985) using the implementation in Maechler et al.
(2024). The MCD estimates for the mean vector and the
variance-covariance matrix are then used to define the reference normal
distribution.

Here we will consider points to be outlying if they are more than
\(5 \sigma\) away from that mean value. With \(p=16\) this corresponds
to a Mahalanobis distance of \(60\) or larger. This will identify \(20\)
out of the \(68\) observations as outlying. Since these are outlying in
different combinations of variables, as found in Mayrhofer and Filzmoser
(2023), we will further separate the outlying points into clusters based
on similarity in direction. The similarity is computed by first
normalizing observations to have length \(1\) and then apply k-means
clustering with Euclidean distance. We use the Dunn index Halkidi,
Batistakis, and Vazirgiannis (2002), computed using Hennig (2024), to
select the preferred number of clusters, \(k=5\).

@@

To compare our results to what is obtained by outlier detection methods,
we reproduce one of the results from Mayrhofer and Filzmoser (2023) in
Figure~\ref{fig-shapley}. On top we can see the cluster assignments for
all points that we have identified as outliers using the
\(p\)-dimensional Mahalanobis distance.

For example we see that cluster 4 contains the years 1962 and 1996, and
for these two years the result from Mayrhofer and Filzmoser (2023) also
indicates unusually low values for the minimum teperature, while there
is disagreement between outlyingness in other variables. Note that
compared to the years 1965 and 1966 they have unusually low minimum
temperature while having average values for the number of days with
maximum temperature above 25°C, and this contrast is also captured in
the projection.

@@

\caption{\label{fig-shapley}Reproduction of the results from Mayrhofer
and Filzmoser (2023), showing which cells have been identified as
outlying by their Shapley Cell Detector algorithm. Red cells indicate
values higher than expected, blue cells values lower than expected. For
years that have been identified as outliers in our approach we show the
cluster assignment at the top.}

@@

The work is related to outlier detection methods (see, Mayrhofer and
Filzmoser 2023, for example), particularly those that use robust
statistics, as illustrated in the weather example. Methods discussed in
Rousseeuw, Raymaekers, and Hubert (2018) and Donoho and Gasko (1992)
describe detecting outlyingness using projections of the data. When
these projections are collected and used as a large set, they define a
high-dimensional confidence region of an indeterminate shape. There is
no analytical definition. However, we could explore the data relative to
the high-dimensional shapes using the same approach described in
Section~\ref{sec-background}. You would generate points on the surface
of the irregular shape, overlay the data on this to visualise with a
tour.

@@

\section*{Supplementary Materials}\label{supplementary-materials}
\addcontentsline{toc}{section}{Supplementary Materials}

@@

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-As85}
Asimov, Daniel. 1985. {``{T}he {G}rand {T}our: {A} {T}ool for {V}iewing
{M}ultidimensional {D}ata.''} \emph{SIAM Journal of Scientific and
Statistical Computing} 6 (1): 128--43.

\bibitem[\citeproctext]{ref-BA86b}
Buja, Andreas, and Daniel Asimov. 1986. {``{G}rand {T}our {M}ethods:
{A}n {O}utline.''} \emph{Computing Science and Statistics} 17: 63--67.

\bibitem[\citeproctext]{ref-BCAH05}
Buja, Andreas, Dianne Cook, Daniel Asimov, and Catherine Hurley. 2005.
{``{C}omputational {M}ethods for {H}igh-{D}imensional {R}otations in
{D}ata {V}isualization.''} In \emph{Handbook of Statistics: Data Mining
and Visualization}, edited by C. R. Rao, E. J. Wegman, and J. L. Solka,
391--414. Amsterdam, The Netherlands: Elsevier/North-Holland.

@@

\bibitem[\citeproctext]{ref-CLBW06}
Cook, Dianne, Eun-Kyung Lee, Andreas Buja, and Hadley Wickham. 2006.
{``{G}rand {T}ours, {P}rojection {P}ursuit {G}uided {T}ours and {M}anual
{C}ontrols.''} In \emph{Handbook of {D}ata {V}isualization}, edited by
C.-H. Chen, W. Härdle, and A. Unwin. Berlin: Springer.

@@

\bibitem[\citeproctext]{ref-FR02}
Fraley, Chris, and Adrian Raftery. 2002. {``Model-Based {C}lustering,
{D}iscriminant {A}nalysis, {D}ensity {E}stimation.''} \emph{Journal of
the American Statistical Association} 97: 611--31.

\bibitem[\citeproctext]{ref-FT74}
Friedman, Jerome H., and John W. Tukey. 1974. {``{A} {P}rojection
{P}ursuit {A}lgorithm for {E}xploratory {D}ata {A}nalysis.''} \emph{IEEE
Transactions on Computing C} 23: 881--89.

\bibitem[\citeproctext]{ref-Ga71}
Gabriel, K. Ruben. 1971. {``The {B}iplot {G}raphical {D}isplay of
{M}atrices with {A}pplications to {P}rincipal {C}omponent {A}nalysis.''}
\emph{Biometrika} 58: 453--67.

\bibitem[\citeproctext]{ref-GH96}
Gower, John C., and David J. Hand. 1996. \emph{Biplots}. London:
Chapman; Hall.

@@

\bibitem[\citeproctext]{ref-Hu85}
Huber, Peter J. 1985. {``{P}rojection {P}ursuit (with Discussion).''}
\emph{Annals of Statistics} 13: 435--525.

\bibitem[\citeproctext]{ref-JS87}
Jones, M. C., and Robin Sibson. 1987. {``{W}hat Is {P}rojection
{P}ursuit? (With Discussion).''} \emph{Journal of the Royal Statistical
Society, Series A} 150: 1--36.

\bibitem[\citeproctext]{ref-lib-med-liver-norms}
Lala, Vasimahmed, Muhammad Zubair, and David A. Minter. 2023. {``Liver
Function Tests.''} \url{https://www.ncbi.nlm.nih.gov/books/NBK482489/}.

@@

\bibitem[\citeproctext]{ref-robustbase}
Maechler, Martin, Peter J. Rousseeuw, Christophe Croux, Valentin
Todorov, Andreas Ruckstuhl, Matias Salibian-Barrera, Tobias Verbeke,
Manuel Koller, Eduardo L. T. Conceicao, and Maria Anna di Palma. 2024.
\emph{Robustbase: Basic Robust Statistics}.
\url{http://robustbase.r-forge.r-project.org/}.

\bibitem[\citeproctext]{ref-mahalanobis}
Mahalanobis, Prasanta Chandra. 1936. {``On the Generalised Distance in
Statistics.''} \emph{Sankhya A} 80 (Suppl 1): 1--7.
\url{https://doi.org/10.1007/s13171-019-00164-5}.

\bibitem[\citeproctext]{ref-MAYRHOFER2023}
Mayrhofer, Marcus, and Peter Filzmoser. 2023. {``Multivariate Outlier
Explanations Using Shapley Values and Mahalanobis Distances.''}
\emph{Econometrics and Statistics}.

@@

\bibitem[\citeproctext]{ref-tourr-cran}
Wickham, Hadley, Dianne Cook, Nicholas Spyrison, Ursula Laa, H. Sherry
Zhang, and Stuart Lee. 2024. \emph{Tour Methods for Multivariate Data
Visualisation}. \url{https://CRAN.R-project.org/package=tourr}.
